---
title: "DEMAND FOR BIKE RENTALS: A report to the management of Capital Bike Share"
author: "Group 3: Carter Che, Brendan Griffin, Dani Herzberg, Nicole Kuker, Anthony Long, Carl Thomas"
date: "2 March 2021"
output: 
  word_document:
    reference_docx: word_styles_reference.docx
---
 
```{r setup, include = FALSE}
# packages
library(tidyverse)
library(lubridate)
library(ggthemes)
# Rmd options
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
# load data
bikes <- read.csv("../data/Capital Bike Sharing data by hour.csv")

# global variables
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
year_levels <- c(2011, 2012)
season_levels <- c("Winter", "Spring", "Summer", "Fall")
weather_levels <- c("Clear", "Mist", "Light Precip.", "Heavy Precip.")
```
\newpage
# Introduction

To be written at later date

# Overall Trends in Bike Rental Demand
```{r Q1 Prep Code, include = FALSE}
# Create for loops for 2011 and 2012
BikeShare_Unique <- unique(bikes$mnth)
BikeShare_Length <- length(unique(bikes$mnth))

  # empty matrix for storage
Month_Store_2011 = matrix(NA, BikeShare_Length , 2)
Month_Table_2011 <- bikes[which(year(bikes$dteday) == 2011) ,]

  # loop for 2011
count = 0
for(ii in BikeShare_Unique){
  count = count + 1
  Month_Store_2011[count, 1] = ii
  X_table <-  Month_Table_2011[which(Month_Table_2011$mnth == ii), ]
  Month_Store_2011[count, 2] = sum(X_table$cnt)
  as.data.frame(Month_Store_2011)
}

  # empty matrix for storage
Month_Store_2012 = matrix(NA, BikeShare_Length , 2)  
Month_Table_2012 <- bikes[which(year(bikes$dteday) == 2012) ,]

  #loop for 2012 
count = 0
for(jj in BikeShare_Unique){
  count = count + 1
  Month_Store_2012[count, 1] = jj+12
  Y_table <-  Month_Table_2012[which(Month_Table_2012$mnth == jj), ]
  Month_Store_2012[count, 2] = sum(Y_table$cnt)
  as.data.frame(Month_Store_2012)
}

Count_per_Month <- data.frame(rbind(Month_Store_2011, Month_Store_2012))

names(Count_per_Month)[1] <- "Month"
names(Count_per_Month)[2] <- "Riders"

qtr_labels <- c("Jan\n2011", "Apr\n2011", "Jul\n2011", "Oct\n2011",
                "Jan\n2012", "Apr\n2012", "Jul\n2012", "Oct\n2012")
```

Question 1 - Anthony & Carl
General trend - positive over the years

```{r}
ggplot(Count_per_Month, aes(x = Month, y = Riders)) + geom_point(alpha = 0.6) +
  # stat_smooth (method = "lm", color = "red", se = FALSE) + 
  coord_cartesian(xlim = c(0.5, 24.5), ylim = c(0, 250000), expand = F) +
  scale_x_continuous(labels = qtr_labels, breaks = seq(1, 24, 3)) + 

  labs(title = "Monthly Bike Rental Demand",
       x = "Month", y = "Total Riders") +
  theme_igray() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.title = element_text(size = 11))
```

# Trends by Season
```{r Q3 Prep Code, include = FALSE}
Seasons <- data.frame(season = c(1,2,3,4), season_levels)
Month <- data.frame(mnth = c(1:12), month_levels)
Years <- data.frame(yr = c(0,1), year_levels)

#Join Season Name to Bike Demand Data Frame
Bike_Season <- bikes
Bike_Season <- left_join(Bike_Season, Seasons, by = "season")
Bike_Season <- left_join(Bike_Season, Years, by = "yr")

#Create Count for Days in Season
Bike_Season$days_cnt <- 1
str(Bike_Season)

#Calculate Daily Demand by Season
Bike_Seasonality <- Bike_Season %>%
  select(Season = season_levels, Year = year_levels, Date = dteday, Demand = cnt, Count = days_cnt) %>%
  group_by(Season, Year) %>%
  summarise(Daily_Demand = sum(Demand)/sum(Count))


Bike_Seasonality$Season <- factor(Bike_Seasonality$Season, levels = season_levels)

# bikes %>% 
#   group_by(yr, season) %>% 
#   summarise(days = length(unique(dteday)),
#             season_demand = sum(cnt),
#             Daily_Demand = season_demand/days,
#             .groups = "drop") %>% 
#   mutate(Year = recode_factor(yr, `0` = "2011", `1` = "2012"),
#          Season = recode_factor(season, levels = !!!season_levels))


#---

#Join Month Name to Bike Demand Data Frame
Bike_Month <- bikes
Bike_Month <- left_join(Bike_Month, Month, by = "mnth")
Bike_Month <- left_join(Bike_Month, Years, by = "yr")

#Create Count for Hours in Month
Bike_Month$hours_cnt <- 1
str(Bike_Month)

#Calculate Hourly Demand by Month
Bike_Monthly <- Bike_Month %>%
  select(Month = month_levels, Year = year_levels, Date = dteday, Demand = cnt, Count = hours_cnt) %>%
  group_by(Month, Year) %>%
  summarise(Hourly_Demand = sum(Demand)/sum(Count), .groups = 'drop') 

Bike_Monthly$Month <- factor(Bike_Monthly$Month, levels = month_levels)
```

Looking at the overall seasonal impact on hourly demand, it is clear that the Winter is a time of depressed demand likely related to the cold temperatures, which will be addressed later. The Spring and Fall are quite similar in demand terms only trailing summer by a small margin on an hourly basis. The seasonal trends in hourly demand are consistent between the two years of data, which indicates it is a continuing trend.

```{r}
ggplot(Bike_Seasonality, aes(x=Season, y=Daily_Demand, fill=Season)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_viridis_d(option = "D", end = 0.75) + 
  facet_wrap(~Year) +
  coord_cartesian(xlim = c(0.5, 4.5), ylim = c(0, 300), expand = F) +
  labs(title = "Average Hourly Demand for Bike Rentals", 
       x = "Season", y = "Hourly Demand",
       fill = "Season") +
  theme_igray() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.title = element_text(size = 11),
        strip.background = element_rect(fill = "grey80"),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8),
        legend.key = element_blank())
```

When the seasons are split into the individual months, there is a more gradual transition into and out of the Winter depressed demand period. The overall trend of seasonality affecting demand holds and January’s demand being 40% of that in June, the peak month. Once again the monthly trend holds consistent between the two years in the dataset. 

```{r}
ggplot(Bike_Monthly, aes(x=Month, y=Hourly_Demand, fill=Month)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_viridis_d(option = "D") + 
  scale_x_discrete(labels = c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")) + 
  facet_wrap(~Year) +
  # coord_cartesian(xlim = c(0.5, 4.5), ylim = c(0, 6900), expand = F) +
  labs(title = "Average Hourly Rental Demand by Month", 
       x = "Month", y = "Hourly Demand",
       fill = "Month") +
  theme_igray() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.title = element_text(size = 11),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8),
        legend.key = element_blank())

```

# Further Trends: Looking at Weather and Rider Type
```{r Q4 Prep Code, include=}
B <- 
  bikes %>% 
  group_by(weathersit) %>% 
  mutate(weathersit = recode_factor(weathersit, !!!weather_levels)) %>%
  summarise(riders = sum(cnt), 
            reg = sum(registered), 
            cas = sum(casual),
            prop_reg = sum(registered) / riders)

b_pivotlong <- 
  bikes %>% 
  select(dteday, hr, weathersit, temp, workingday, casual:cnt) %>%
  pivot_longer(cols = c("casual", "registered"), names_to = "ridertype", values_to = "numRiders") %>%
  mutate(weathersit = recode_factor(weathersit, !!!weather_levels)) %>%
  group_by(weathersit)

b_pivotlong.summary <- 
  b_pivotlong %>% 
  group_by(weathersit, ridertype) %>% 
  summarise(N = sum(numRiders)) %>% 
  mutate(prop_ridergivenweather = N/sum(N))
b_pivotlong.summary <- 
  b_pivotlong.summary %>% 
  group_by(ridertype) %>% 
  mutate(prop_weathergivenrider = N/sum(N))
```

Hourly weather conditions were documented within the dataset and categorized as Clear, Mist, Light Precipitation, and Heavy Precipitation. As weather conditions worsen, there is a steep decrease in overall ridership. Though there are more registered riders than casual riders, this decrease in ridership holds true when looking at both subsets of rider type.

```{r}
# Total Riders by Weather Type
ggplot(B, aes(x = weathersit, y = riders, fill = weathersit)) + 
  geom_col() + 
  geom_text(aes(label = riders), vjust = -0.5) + 
  scale_fill_viridis_d(option = "D") + 
  coord_cartesian(xlim = c(0.5, 4.5), ylim = c(0, 2500000), expand = F) +
  labs(title = "Total Riders by Weather Type", 
       x = "Weather Type", y = "Total Riders",
       fill = "Weather Type") +
  theme_igray() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.title = element_text(size = 11),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8),
        legend.key = element_blank())
```

As weather conditions worsen, casual riders make up a decreasing proportion of total ridership. This suggests that casual riders may be more influenced by the weather, and are more likely to choose other modes of transportation when weather conditions are unpleasant or uncomfortable.

```{r}
ggplot(b_pivotlong.summary, aes(weathersit, y = N, fill = ridertype)) + 
  geom_bar(stat = "identity", position = "fill") + 
  labs(title = "How Weather Affects Ridership", 
       x = "Type of Weather", y = "Proportion of Rider Type",
       fill = "Rider Type") + 
  geom_text(data = subset(b_pivotlong.summary, ridertype == "casual"),
            aes(label = round(prop_ridergivenweather, 3)), position = "fill", vjust = -0.2) + 
  geom_text(data = subset(b_pivotlong.summary, ridertype == "registered"),
            aes(label = round(prop_ridergivenweather, 3)), color = "grey90", position = "fill", vjust = 14) + 
  scale_fill_viridis_d(option = "D", begin = 0, end = .45, direction = -1) + 
  theme_igray() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.title = element_text(size = 11),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 8),
        legend.key = element_blank())
```

# Predicting Total Bike Rental Demand
Question 5 - Fit a linear model predicting the total bike rental demand from daily temperature. What kind of insights can you generate? (make sure to write the linear model and interpret it in the context of the data)

### Daily Temperature
```{r Q5 Prep Code, include = FALSE}
Temp_Plot <- bikes %>%
  ggplot(aes(x = temp,
             y = cnt)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color="grey60") 

Temp_Plot

Predict <- lm(cnt ~ temp, data = bikes)
summary(Predict)
coefficients(Predict)
```
The equation of this linear model is: 
    Expected Demand = -0.04 + (381. 30 * Temp)
    (where temperature is the normalized temperature in the data set)

The linear model is statistically significant given the p-value is below the alpha of 0.05 (it is also below an alpha of 0.01). The coefficient t-value is a measure of how many standard deviations our coefficient estimate is far away from 0. In the temperature linear model, the t-statistic value is relatively far away from zero and is large relative to the standard error, which could indicate a relationship exists.

However, the R squared value is only 0.1638, which indicates that the linear model is not a strong fit to the data and that the relationship between temperature and demand is not particularly strong. This R squared means roughly 16.4% of the variance found in overall demand can be explained by the actual temperature.

The scatterplot indicates that there is a wide range of demand at almost all temperatures. The demand range only tightens at the extremes of the temperature scale. 

### Daily Feeling Temperature
```{r Q6 Prep Code, include = FALSE}
Atemp_Plot <- bikes %>%
  ggplot(aes(x = atemp,
             y = cnt)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color="grey60") 

Atemp_Plot

Predict_A <- lm(cnt ~ atemp, data = bikes)
summary(Predict_A)
coefficients(Predict_A)
```

Question 6 - Fit another linear model predicting total daily bike rentals from daily feeling temperature. Write the linear model, interpret the slope, etc.

The equation of this linear model is: 
    Expected Demand = -11.88 + (423.18 * Feels Like Temp)
    (where feels like temperature is the normalized feels like temperature in the data set)

The linear model is statistically significant given the p-value is below the alpha of 0.05 (it is also below an alpha of 0.01). The p-value tests the null hypothesis that there is no correlation between the feels like temperature and overall demand and we can reject the null hypothesis.

However, the R squared value is only 0.1607, which indicates that the linear model is not a strong fit to the data and that the relationship between the feels like temperature and demand is not particularly strong. Only 16.1% of the variance in the overall demand can be explained by the feels like temperature. 

The slope of the linear model is steeper than in the actual temperature model indicating a one degree change in the feels like temperature has a bigger impact on demand than a one degree change in the actual temperature. 

The scatterplot indicates that there is a wide range of demand at almost all feels like temperatures. The demand range only tightens at the extremes of the temperature scale. 


### Which is the better predictor?

Part of Question 6: Is the temperature or feeling temperature a better predictor of bike rentals?

The actual temperature is a better predictor of overall demand for bike rentals. The higher R squared and lower standard error in the temperature linear model indicates a marginally better fit than the feels like temperature model. That being said, the R squared is low for both models and the standard error high. Neither model is a particularly good fit for the data and does not explain a large portion in the demand for bike rentals.

# Correlation between Rental Demand and Typical Office Hours
```{r Q2 Prep Code, include = FALSE}
#Create map of hourly demand to time of day windows
TOD <- read.csv("data/TOD.csv") %>% 
  as_tibble()
names(TOD)[names(TOD) == "ï..Hour"] <- "Hour"

#Create clean dataset for hourly demand
Hypothesis <- select(bikes, Hour = hr, Weekday = weekday, Holiday = holiday, Workingday = workingday, 
         Casual = casual, Registered = registered, Total = cnt) 

Hypothesis <- left_join(Hypothesis, TOD, by = "Hour")

#Create subset of hourly demand for day-time
Day_Demand <- Hypothesis %>%
  filter(Window == "Day")

#Create mean of hourly demand for day
Avg_DayDemand <- mean(Day_Demand$Total)

#Create subset of hourly demand for office peak
Office <- Hypothesis %>%
  filter(TOD == "Office")

#Null hypothesis: Office Peak Demand >= Total Day Demand

#Test whether office peak demand is greater than total demand
t.test(Office$Total, mu=Avg_DayDemand, alternative="less")

#Conclusion: accept the null as p-value is greater than alpha 0.05

#Create subset of hourly demand for night
Night <- Hypothesis %>%
  filter(TOD == "Night")

#Null hypothesis: Night Demand <= Total Demand

#Test whether night demand is less than total demand
t.test(Night$Total, mu=Avg_DayDemand, alternative="greater")

#Conclusion: accept the null as p-value is greater than alpha 0.05
```

### Is there a "rush hour" for bike rentals?

Hypothesis: There must be high demand during office timings. Early morning and late evening can have different trends (cyclist) and low demand from 10:00pm to 4:00am. Do you agree?

**Add distrubtion of demand over hours to final report

To assess whether there is higher demand during office hours than during the overall daytime hours, a hypothesis test was run. The null hypothesis being tested was stated as "Office Hour Demand" (or the sample mean) is greater than or equal to the "Daytime Demand" (or the population mean). The result of the hypothesis test was that we did not reject the null hypothesis and we have evidence that "Office Hour Demand" is greater than the "Daytime Demand". We ran this same test for "Office Hour Demand" being greater than or equal to the "Trough Daytime Demand" (or the hours outside office timings) and, again, did not reject the null hypothesis and there is evidence that Office Hours have more demand than the remaining Daytime Hours.

**Need to do early morning versus late evening?

The overnight hours (20:00 to 04:00) had a hypothesis that their demand was lower than the Daytime Demand and so another hypothesis test was run. The null hypothesis was stated as "Overnight Demand" is less than or equal to "Daytime Demand". Again, the result was to not reject the null hypothesis as the test statistic was not in the rejection range. So, there is evidence that the overnight hours are low demand periods.

### Workdays vs. Weekends/Holidays

Hypothesis: Registered users demand more bike on weekdays compared to the weekend or holiday. Do you agree?

# Recommendations and Conclusions

Question 7 - What is your overall recommendation about the demand for bike rentals for the Capital Bike Share management? Share your insights with the company based on your analysis.

* Overnight discounts or peak commute surcharges?
* Lower staffing and inventory in winter? Good time for maintenance?
* 

\newpage
# Appendix
